\name: Combining Bottom-up and top-down approaches through graph learning over interaction networks for drug-target-interaction prediction

### 2021-03-21 paper/main.tex
.

We trained, validated and finally tested all considered models on the
STITCH dataset using a 5-fold cross-validation over a protein split;
we then selected the best-performing models (with respect to
\textit{MacroAUC} and $MicroAUC_p$, see Section \ref{sec:evaluation})
and evaluate them in a 5-fold protein-split cross-validation on the
Yamanishi benchmark dataset to avoid validation overfitting and yield
more realistic testing results. To evaluate the influence of the
different features separately, and to determine whether they
``localize'' on the PPI graph (and therefore can be exploited
successfully by the graph neural networks), we train and evaluate
models with different types of features, and with and without
inclusion of the PPI graph, separately. We comparing the molecular
(MolPred) and phenotype-based (OntoPred) prediction model, and a
combination of both when concatenating both types of features.  Table
\ref{tab:Results} (a)+(b) shows the results of these experiments.

We find that the model using ontology-based features
(\textit{OntoPred}) is showing better performance on STITCH compared
to using only molecular features. We also observe that only the model
using ontology-based features results in increased performance when
incorporating the PPI graph. This increase can be observed with
different graph neural network architectures and configurations. While
the {\em GCNConv} and {\em GENConv} architecture already shows some
minor improvement, the \textit{ResGraphConv} blocks add a large amount
of additional learnable parameters to the network, leading to more
expressive power. To test whether the improvement is due to the number
of learnable parameters added or the result of better exploiting the
information about PPIs, we experiment with a graph model in which all
graph convolutional neural layers in the residual blocks are removed,
resulting in a model with similar parameters but without the ability
to use graph-based information. This pruned network, with no
information on the protein--protein interactions, reached very similar
results to the original \textit{OntoPred} model and showed no
improvement.

The improvements when including the graph are only provided by the
\textit{GENConv} graph convolution scheme which includes the
ResGraphConv blocks; \textit{GCNConv} and other graph
convolutional methods fail to achieve any gain in comparison to the
plain \textit{OntoPred} performance even when combined with the
residual blocks. The discrepancy between {\em GENConv} and other graph
convolutional methods may be the result of numerical instability and
fading signals \cite{}.
Nevertheless, our results demonstrate that the inclusion of graph
information increases performance when ontology-based features are
used, which allows us to conclude that information about protein
functions localizes on the graph whereas molecular features do not.

@@

\subsection{Protein-centric evaluation}
The goal of \name is to find candidate drugs that target a specific
protein; however, so far, we do not evaluate this application
but rather how \name would perform in finding plausible drug--target
interactions among all possible interactions. This evaluation does not
correspond to the application of \name in finding drugs that target a
specific protein.
To provide a better
estimate on how \name performs for individual protein targets, we use
micro-averages between proteins and compute the \textit{MicroAUC} (see
Section \ref{sec:evaluation}); micro-averages average the performance
(true and false positive rates) per protein instead of across all
drug--protein pairs, and can therefore better estimate how \name
performs when tasked with finding a drug that targets a specific
protein.

>>>>

We trained, validated and finally tested all considered models on the
STITCH dataset using a 5-fold cross-validation over a protein split;
we then selected the best-performing models (with respect to
\textit{MacroAUC} and $MicroAUC_p$\todo{which is it, macro or micro?},
see Section \ref{sec:evaluation}) and evaluate them in a 5-fold
protein-split cross-validation on the Yamanishi benchmark dataset to
avoid validation overfitting and yield more realistic testing
results. To evaluate the influence of the different features
separately, and to determine whether they ``localize'' on the PPI
graph (and therefore can be exploited successfully by the graph neural
networks), we train and evaluate models with different types of
features, and with and without inclusion of the PPI graph,
separately. We comparing the molecular (MolPred) and phenotype-based
(OntoPred) prediction model, and a combination of both where we
concatenate both types of features.  Table
\ref{tab:Results}\todo{check -- not currently in the table} shows the
results of these experiments.

We find that the model using ontology-based features
(\textit{OntoPred}) is showing better performance on STITCH compared
to using only molecular features. We also observe that only the model
using ontology-based features results in increased performance when
incorporating the PPI graph. This increase can be observed with
different graph neural network architectures and configurations. While
the {\em GCNConv} and {\em GENConv} architecture already shows some
minor improvement, the use of \textit{ResGraphConv} results in larger
performance improvements.  {\em ResGraphConv} blocks add a large
amount of additional learnable parameters to the network, leading to
more expressive power. To test whether the observed improvement is due
to the number of learnable parameters added or the result of better
exploiting the information about PPIs, we experiment with a graph
model in which all graph convolutional neural layers in the residual
blocks are removed, resulting in a model with similar parameters but
without the ability to use graph-based information. This pruned
network, with no information on the protein--protein interactions,
reached very similar results to the original \textit{OntoPred} model
and showed no improvement.

The improvements when including the graph are only provided by the
\textit{GENConv} graph convolution scheme which includes the
{\em ResGraphConv} blocks; \textit{GCNConv} and other graph
convolutional methods fail to achieve any gain in comparison to the
plain \textit{OntoPred} performance even when combined with the
residual blocks. The discrepancy between {\em GENConv} and other graph
convolutional methods may be the result of numerical instability and
fading signals \cite{}.

Our results demonstrate that the inclusion of graph information
can increase performance when ontology-based features are used but not
when molecular features are used alone. This observation allows us to
conclude that information about protein functions localizes on the
graph whereas molecular features do not.

@@

\subsection{Protein-centric evaluation}
The goal of \name is to find candidate drugs that target a specific
protein; however, so far, we do not evaluate this application but
rather how \name would perform in finding plausible drug--target
interactions among all possible interactions (since we use the
MacroAUC as our main evaluation measure). This evaluation does not
correspond to the application of \name in finding drugs that target a
specific protein.  To provide a better estimate on how \name performs
for individual protein targets, we use micro-averages between proteins
and compute the \textit{MicroAUC} (see Section \ref{sec:evaluation});
to determine {\em MicroAUC}, we average the performance (true and
false positive rates) per protein instead of across all drug--protein
pairs; the resulting measure can therefore better estimate how \name
performs when tasked with finding a drug that targets a specific
protein.
